---
title: "Improve generation of the null distribution"
output: html_document
---

```{r}
library(ape)
```


```{r, include = FALSE}
plot_tree_w_color_edges <- function(tr,
                                  edges_to_highlight,
                                  geno_confidence,
                                  edge_color_na = "grey",
                                  edge_color_bright = "red",
                                  title = "title", 
                                  legend_bright = "highligted edge", 
                                  legend_baseline = "nonhighlighted edge"){
  # Function -------------------------------------------------------------------
  tree_legend_cex <- 0.5
  edge_color_baseline <- "black"
  edge_color <- rep(edge_color_baseline, ape::Nedge(tr))
  edge_color[edges_to_highlight == 1] <- edge_color_bright
  edge_color[geno_confidence == 0] <- edge_color_na # grey out long

  edge_width <- 1
  if (ape::Ntip(tr) >= 200) { edge_width <- 0.25 }

  graphics::par(mar = c(4, 4, 4, 4))
  graphics::plot(tr,
                 font = 1,
                 edge.color = edge_color,
                 main = title,
                 label.offset = 0.25,
                 adj = 0,
                 cex = tree_legend_cex,
                 edge.width = edge_width)
  
  graphics::legend("topleft",
                   bty = "o",
                   legend = c(legend_baseline,
                              legend_bright,
                              "Low confidence"),
                   col = c(edge_color_baseline,
                           edge_color_bright,
                           edge_color_na),
                   lty = 1,
                   ncol = 1,
                   lwd = 1,
                   cex = tree_legend_cex)
}
```

Generate example data: tree, genotype, genotype transitions, and phenotype 
transition. 
```{r}
# set up a tree
num_samples <- 6
set.seed(1)
temp_tree <- ape::rtree(num_samples)
temp_tree$node.labels <- rep(100, ape::Nnode(temp_tree))
num_edge <- ape::Nedge(temp_tree)

# set up genotype transitions per edge: a list with a vector for each genotype
num_genotypes <- 15
temp_geno_trans <- temp_hi_conf_edges <- rep(list(0), num_genotypes)
set.seed(1)
for (k in 1:num_genotypes) {
  temp_geno_trans[[k]] <- c(0, 0, 0, 1, 0, 0, 0, 1, 0, 0)
  temp_hi_conf_edges[[k]] <- rep(1, num_edge)
  temp_hi_conf_edges[[k]][sample(1:num_edge, 3, replace = TRUE)] <- 0 
}
temp_geno_trans[[14]] <- c(0, 1, 1, 1, 0, 0, 1, 1, 0, 0)
temp_geno_trans[[15]] <- c(0, 0, 1, 1, 0, 0, 0, 1, 0, 0)

# set up a vector with phenotype transition per edge
temp_pheno_trans <- c(1, 1, 1, 1, 0, 0, 0, 0, 0, 0)

# set up a "fake" genotype that doesn't match recon or transition because it 
# doesn't need to match for this example
temp_geno <- matrix(1, ncol = num_genotypes, nrow = ape::Ntip(temp_tree))

# set up number of permutations
temp_perm <- 300

temp_num_edges_with_geno_trans <- sapply(temp_geno_trans, function(x) sum(x))
temp_num_hi_conf_edges <- sapply(temp_hi_conf_edges, function(x) sum(x))
```

# CURRENT VERSION 
Run through the code for one genotype (the index is the genotype number)
```{r}
index <- 2
permuted_geno_trans_mat <-
  matrix(nrow = temp_perm, ncol = temp_num_edges_with_geno_trans[index])
redistributed_hits <- matrix(0, nrow = temp_perm, ncol = ape::Nedge(temp_tree))

# set j
for (j in 1:temp_perm) {
permuted_geno_trans_mat[j, ] <-
  sample(1:temp_num_hi_conf_edges[index],
         size = temp_num_edges_with_geno_trans[index],
         replace = FALSE,
         prob = temp_tree$edge.length[as.logical(temp_hi_conf_edges[[index]])] /
           sum(temp_tree$edge.length[temp_hi_conf_edges[[index]]]))
}

for (m in 1:nrow(permuted_geno_trans_mat)) {
  redistributed_hits[m, ][permuted_geno_trans_mat[m, ]] <- 1
}

CURRENT_VERSION_redistributed_hits <- redistributed_hits
CURRENT_VERSION_permuted_geno_trans_mat <- permuted_geno_trans_mat

```

# PROPOSED NEW VERSION

```{r}
# convert example code above (originally adapted from unit test) to 
num_perm <- temp_perm
high_conf_edges <- temp_hi_conf_edges
number_edges_with_geno_trans <- temp_num_edges_with_geno_trans
tr <- temp_tree
# --- 

permuted_geno_trans_mat <-
  matrix(nrow = temp_perm, ncol = temp_num_edges_with_geno_trans[index])
redistributed_hits <- matrix(0, nrow = temp_perm, ncol = ape::Nedge(temp_tree))

for (j in 1:num_perm) {
    permuted_geno_trans_mat[j, ] <-
      sample(which(as.logical(high_conf_edges[[index]])),
             size = number_edges_with_geno_trans[index],
             replace = FALSE,
             prob = tr$edge.length[as.logical(high_conf_edges[[index]])] /
               sum(tr$edge.length[as.logical(high_conf_edges[[index]])]))
}

for (m in 1:nrow(permuted_geno_trans_mat)) {
  redistributed_hits[m, ][permuted_geno_trans_mat[m, ]] <- 1
}

PROPOSED_NEW_VERSION_redistributed_hits <- redistributed_hits
PROPOSED_NEW_VERSION_permuted_geno_trans_mat <- permuted_geno_trans_mat
```


```{r}
prob <- tr$edge.length[as.logical(high_conf_edges[[index]])] /
               sum(tr$edge.length[as.logical(high_conf_edges[[index]])])

all_prob <- rep(0, num_edge)
all_prob[which(as.logical(high_conf_edges[[index]]))] <- prob

expected_count <- 
  all_prob * sum(PROPOSED_NEW_VERSION_permuted_geno_trans_mat != 0)
```

# Plot probability of selecting each edge
This probability is the same for both the current and proposed new version of
the code. 

The probability for each edge is the relative edge length; all low confidence
edges are removed from consideration. 

```{r}
plot(x = 1:10, 
       y = all_prob, 
       cex = 2,
          xaxt = "n",
     xlab = "Edge", 
     ylab = "Probability",
     main = "Probability of selecting edge \nBased on length & considering reconstruction confidence",
     col = "orange")
axis(1, at = 1:10)

```

# Now let's compare which edges get chosen in the two versions

## CURRENT VERSION 

```{r}
plot(table(CURRENT_VERSION_permuted_geno_trans_mat), 
     xlim = c(0, 10),
          ylim = c(0, 140),
     xlab = "Edge number", 
     ylab = "Count",
     col = "red", 
          xaxt = "n",
     main = "Edges chosen by CURRENT VERSION;\n black x=lo conf; grey caret=hi conf")
points(which(!as.logical(high_conf_edges[[index]])),  c(0, 0, 0), 
     xlim = c(0, num_edge),
     col = "black", pch = "x", cex = 3)
points(which(as.logical(high_conf_edges[[index]])),  c(0, 0, 0, 0, 0, 0, 0), 
     xlim = c(0, num_edge),
     col = "grey", pch = "^", cex = 3)
points(x = 1:10, 
       y = expected_count, 
       col = "orange", 
       cex = 2)
axis(1, at = 1:10)

```

In this particular example: 7/10 edges are high confidence and the sampling 
function picks `r sum(number_edges_with_geno_trans[index])` edges during each
permutation. As currently written, the null distribution is made up from only 
edges 1:7. This is skewing the null distribution based on the default edge order
(I think the ape default) AND is going to make P-values artificially smaller if
- as in this case - there are an enriched number of low confidence edges earlier
in the tree order. 

## PROPOSED NEW VERSION 
```{r}
plot(table(PROPOSED_NEW_VERSION_permuted_geno_trans_mat), 
     xlim = c(0, 10),
     ylim = c(0, 140),
     xlab = "Edge number", 
     ylab = "Count",
     col = "blue", 
          xaxt = "n",

     main = "Edges chosen by PROPOSED NEW VERSION;\n black x=lo conf; grey caret=hi conf")
points(which(!as.logical(high_conf_edges[[index]])),  c(0, 0, 0), 
     xlim = c(0, num_edge),
     col = "black", pch = "x", cex = 3)
points(which(as.logical(high_conf_edges[[index]])),  c(0, 0, 0, 0, 0, 0, 0), 
     xlim = c(0, num_edge),
     col = "grey", pch = "^", cex = 3)
axis(1, at = 1:10)


points(x = 1:10, 
       y = expected_count, 
       col = "orange", 
       cex = 2)
```

In this particular example: 7/10 edges are high confidence and the sampling 
function picks `r sum(number_edges_with_geno_trans[index])` edges during each
permutation. I propose that the new null distribution be made up from only 
the high confidence edges (3, 4, 5, 6, 8, 9, & 10). This generates a null 
distribution that mimics far more closely the relative tree edge lengths. This
is what we described the algorithm as doing in the paper. 

# Conclusions
Currently, the plan is to implement the "Proposed New Version" pending a few 
discussions, unit testing, etc... 

